---
title: "Group 4 - COVID Data Presentation"
output: 
  flexdashboard::flex_dashboard:
    orientation: column
    theme: sandstone
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```


```{r load libraries}
library(tidyverse)
library(dplyr)
library(magrittr)
library(forcats)
library(ggplot2)
library(ggthemes)
library(readr)
library(lubridate)
library(tidyverse)
library(tigris)
library(viridis)
library(ggthemes)
library(plotly)
library(stringr)
```

Sidebar {.sidebar}
=====================================

**Data Sources:**

Two datasets were used to create this dashboard.  
  
New York Times: [here](https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-counties.csv)  
  
Colorado Department of Public Health and Environment (CDPHE) COVID-19 County Level Open Data Repository: [here](https://data-cdphe.opendata.arcgis.com/datasets/cdphe-covid19-county-level-open-data-repository)  
  
The New York Times database was used to obtain number of COVID-19 cases and deaths by day per Colorado county and to find daily new cases and deaths. The CDPHE database was also used to data on COVID-19 rates per 100,000 by county and hospitalization data by county.  
  
**Maps**  
  
*COVID Cases by County Classification:* Three maps are presented that organize counties by frontier, rural, and urban. A color-coded scale is provided that indicates COVID-19 cases per 100,000. Hold your cursor over each county to obtain county name and COVID-19 rate per 100,000 people. Rates are for November 11th, 2020.  
  
**Epi-Plots**  
  
*COVID Cases, Deaths, and Hospitalizations:* This figure depicts Colorado COVID-19 new cases, new deaths, and current hospitalizations over time. The red line corresponds to cases, green to deaths, and blue to current hospitalizations. The curve indicates Fourth of July and Halloween timepoints to highlight differences in cases, deaths, and hospitalizations around these times characterized by increased social interaction. Use this figure to search for temporal associations between time points of interest and trends in COVID-19 cases, deaths, and hospitalizations.  
  
*Total recorded cases by county:* This interactive tool allows the user to determine total cases over time in each Colorado county. Line colors correspond to frontier (red), rural (green), or urban (blue). Hold the cursor at any point along a line to see county name and total number of COVID-19 cases up that date. This interactive figure allows users to see associations between frontier, rural, and urban settings and total COVID-19 cases.  
  
*Rate per 100000 for 3 counties per class.* This interactive tool provides COVID-19 rate (per 100,000) over time for three counties from each of the three classes. Red lines correspond to frontier counties. Green lines correspond to rural counties. Blue lines correspond to urban settings. Hold the cursor over a line at any point to obtain COVID-19 rate (per 100,000) for that date, county name, and class. This tool allows the user to compare daily COVID-19 rates by class and county

MAPS
=====================================
    
Column
-------------------------------------
    
### COVID Cases by County Classification
    
```{r message = FALSE, warning = FALSE, include = FALSE}
co_counties <- counties(state = 08, cb = FALSE, resolution = "20m") %>% #read in SF files for counties
  rename(County = NAME)

```

```{r}
cdphe <- read_csv("Data/Colorado_COVID-19_Positive_Cases_and_Rates_of_Infection_by_County_of_Identification.csv") %>% #CDPHE data for rates
  rename(County = LABEL) %>%
  slice(1:64)

covid <- read_csv("https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-counties.csv") %>%
  filter(state == "Colorado") #read in NYT covid data

#covid_map <- full_join(co_counties, covid, by = "County") %>% #join NYT data to SF
  #filter(date == "2020-12-01") #filter for 12-01-2020

county_bins <- tibble(County = c("Larimer", "Mesa", "Weld", "Boulder",
                                 "Broomfield", "Gilpin", "Adams",
                                 "Clear Creek", "Denver", "Arapahoe",
                                 "Jefferson", "Douglas", "Elbert", "Park",
                                 "Teller", "El Paso", "Pueblo",
                                 "Logan", "Phillips", "Morgan", "Prowers", 
                                 "Crowley","Otero", "Grand", "Routt", "Eagle", 
                                 "Summit","Garfield", "Pitkin", "Lake",
                                 "Chaffee","Fremont", "Delta", "Montrose",
                                 "Ouray", "Alamosa","Rio Grande", "Conejos", 
                                 "Archuleta", "La Plata","Montezuma","Moffat", 
                                 "Rio Blanco", "Jackson","Gunnison",
                                 "San Miguel", "Dolores", "San Juan",
                                 "Hinsdale","Mineral", "Saguache", "Custer", 
                                 "Huerfano","Costilla", "Las Animas", "Baca",
                                 "Bent","Kiowa", "Lincoln", "Cheyenne",
                                 "Kit Carson","Yuma", "Washington", "Sedgwick"),
                      class = c("urban","urban","urban","urban","urban","urban",
                                "urban","urban","urban","urban","urban","urban",
                                "urban","urban","urban","urban","urban", "rural",
                                "rural","rural","rural","rural","rural","rural",
                                "rural","rural","rural","rural","rural","rural",
                                "rural","rural","rural","rural","rural","rural",
                                "rural","rural","rural","rural","rural", 
                                "frontier","frontier","frontier","frontier",
                                "frontier","frontier","frontier","frontier",
                                "frontier","frontier","frontier","frontier",
                                "frontier","frontier","frontier","frontier",
                                "frontier","frontier","frontier","frontier",
                                "frontier","frontier","frontier")) %>%
  mutate(Class = str_to_title(class))

rate_map <- full_join(co_counties, cdphe, by = "County") #join cdphe data to SF

cdphe_class <- full_join(rate_map, county_bins, by = "County") %>%
  rename(Rate = County_Rate_Per_100_000)

# ggplot() + #covid map by raw cases
#  geom_sf(data = co_counties) +
#  geom_sf(data = covid_map, aes(fill = cases)) +
#  scale_fill_viridis()

map <- ggplot() + #covid map by rates 
  geom_sf(data = co_counties) +
  geom_sf(data = cdphe_class, aes(fill = Rate, group = County)) +
  facet_wrap(~Class) +
  scale_fill_viridis() +
  theme_map() +
  theme(legend.position="bottom")

anmap <- ggplotly(map)
anmap

```


Epi-plots {data-orientation=columns}
=====================================  

Column
-------------------------------------
   
### COVID Cases, Deaths, and Hospitalizations

```{r}
# Read in county case/deaths NY times data

counties <- read_csv("https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-counties.csv") %>%
  filter(state == "Colorado")

# use lag() to change values from total # of cases/deaths to daily new cases/deaths:

county_new <- counties %>%
  group_by(county) %>%
  arrange(date) %>%
  mutate(new_cases = cases - lag(cases, default = first(cases))) %>%
  mutate(new_deaths = deaths - lag(deaths, default = first(deaths)))

# join with Elizabeth's county classification dataframe for use in epi graphs:
# Note: this is Elizabeth's code.

county_bins <- tibble(county = c("Larimer", "Mesa", "Weld", "Boulder",
                                 "Broomfield", "Gilpin", "Adams",
                                 "Clear Creek", "Denver", "Arapahoe",
                                 "Jefferson", "Douglas", "Elbert", "Park",
                                 "Teller", "El Paso", "Pueblo",
                                 "Logan", "Phillips", "Morgan", "Prowers", "Crowley",
                                 "Otero", "Grand", "Routt", "Eagle", "Summit",
                                 "Garfield", "Pitkin", "Lake", "Chaffee",
                                 "Fremont", "Delta", "Montrose", "Ouray", "Alamosa",
                                 "Rio Grande", "Conejos", "Archuleta", "La Plata",
                                 "Montezuma","Moffat", "Rio Blanco", "Jackson", 
                                 "Gunnison","San Miguel", "Dolores", "San Juan",
                                 "Hinsdale","Mineral", "Saguache", "Custer", "Huerfano",
                                 "Costilla", "Las Animas", "Baca", "Bent",
                                 "Kiowa", "Lincoln", "Cheyenne", "Kit Carson",
                                 "Yuma", "Washington", "Sedgwick"),
                      class = c("urban","urban","urban","urban","urban","urban",
                                "urban","urban","urban","urban","urban","urban",
                                "urban","urban","urban","urban","urban", "rural",
                                "rural","rural","rural","rural","rural","rural",
                                "rural","rural","rural","rural","rural","rural",
                                "rural","rural","rural","rural","rural","rural",
                                "rural","rural","rural","rural","rural", 
                                "frontier","frontier","frontier","frontier",
                                "frontier","frontier","frontier","frontier",
                                "frontier","frontier","frontier","frontier",
                                "frontier","frontier","frontier","frontier",
                                "frontier","frontier","frontier","frontier",
                                "frontier","frontier","frontier"))

county_class <- full_join(county_new, county_bins, by = "county")

# consolidate data from county info to state info using aggregate():

co_cases <- aggregate(county_new["new_cases"], by = county_new["date"], sum)
co_deaths <- aggregate(county_new["new_deaths"], by = county_new["date"], sum)

# Read in hospitalization data for CO:

hosp_data <- read_csv("data/covid19_hospital_data_2020-11-29.csv")

# filter hospital data down to current hospitalization of confirmed covid cases. 

hosp_data_conf <- hosp_data %>%
  filter(description == "Currently Hospitalized" & metric == "Confirmed COVID-19")

# join dataframes (automatically joins by date for "full_join()")

co_cases_deaths <- full_join(co_cases, co_deaths)
co_full <- inner_join(co_cases_deaths, hosp_data_conf, by = "date") %>%
  select(1, 2, 3, 9) %>%
  rename(new_hospitalizations = "value") %>% 
  rename(Cases = "new_cases") %>% 
  rename(Deaths = "new_deaths") %>% 
  rename(Hospitalizations = "new_hospitalizations")

#Plot of cases v deaths v hospitalizations (using aggregated datasets)

g4_epi_plot_1 <- co_full %>%
  ggplot() +
  geom_line(aes(x = date, y = Cases, color = "Cases")) +
  geom_line(aes(x = date, y = Deaths, color = "Deaths")) +
  geom_line(aes(x = date, y = Hospitalizations, color = "Hospitalizations")) +
  geom_vline(aes(xintercept = ymd("2020-07-04")), linetype = "dashed",
             alpha = 0.5) +
  geom_vline(aes(xintercept = ymd("2020-10-31")), linetype = "dashed",
             alpha = 0.5) +
  geom_text(aes(x = ymd("2020-07-04"), label="Fourth of July", y=4000), color = "grey36",
            hjust = 1.1) +
  geom_text(aes(x = ymd("2020-10-31"), label="Halloween", y=5000), color = "grey36",
            hjust = 1.1) +
  labs(x = "Date", y = "Cases, Deaths, and Hospitalizations", color = "Legend") +
  theme_classic() 

g4_epi_plot_1
```   
 
 
Column
-------------------------------------

### Total recorded cases by county
    
```{r}
#Plot of total recorded cases by county (use county_class data so you can color by class)
#color by population category instead of by county to match Elizabeth's graphs. 

g4_epi_plot_2 <- county_class %>%
  na.omit %>%
  ggplot() +
  geom_line(aes(x = date, y = cases, group = county, col = class)) +
  labs(x = "Date", y = "Total Cases") +
  theme_classic()

# wrap plot in ggplotly() to animate:

library(plotly)
library(htmlwidgets)

g4_epi_plot_2_anm <- ggplotly(g4_epi_plot_2)
g4_epi_plot_2_anm
```

### Rate per 100000 for 3 counties per class

```{r}
#Plot of rate per 10000 by county (note this only works if loading initial chunks of code 
#and assigning all objects in flexdashboard)

county_class_rates <- county_class %>%
  mutate(rate_per_100000 = new_cases/100000) %>% 
  filter(county == c("Lincoln", "Bent", "Kit Carson", "La Plata", "Logan", 
                     "Crowley", "El Paso", "Weld", "Larimer"))
#Lincoln, Bent, Kit Carson
#La Plata, Logan, Crowley
#El Paso, Weld, Larimer

g4_epi_plot_3 <- county_class_rates %>%
  na.omit() %>%
  ggplot() +
  geom_line(aes(x = date, y = rate_per_100000, group = county, col = class)) +
  labs(x = "Date", y = "Rate per 100000") +
  theme_classic()

g4_epi_plot_3_anm <- ggplotly(g4_epi_plot_3)

g4_epi_plot_3_anm

```

